### 2010-02-26 trunk/deliverables/D2_2_Ontology/fast_d2_2_2_ontology.tex
the final version that will be submitted to the EC. Mainly revised ontology code, introduction and conclusion
git-svn-id: https://svn.forge.morfeo-project.org/fast-fp7project@603 d8b0c59b-6544-0410-94d9-83a5658a058e

\lstdefinelanguage{turtle} 
{morekeywords={@prefix, a}, 
sensitive=false, 
morecomment=[l]{\#}, 
morestring=[b]",
}

@@

This deliverable defines the ontology and conceptual model for the semantic characterisation of complex gadgets in the FAST project. As such, it feeds directly into the various implementation efforts within the project. In the document, we present the ontology development methodology (Methontology) and design principles we have applied. Following Methontology, our design and development process results in a number of different living documents, which both function as a development tool, as well as documentation for the ontology itself. The major part of the deliverable is made up of these documents, resulting in the final \emph{implementation document} which formally defines the ontology in OWL DL semantics.

@@

\begin{tabular}
	{| >{\columncolor{fast@lightgrey}}p{3.25cm}|p{1.4cm}|p{3.28cm}|p{1.6cm}|p{3.29cm}|}
	\hline
	\textcolor{white}{\textbf{Delivery data}} & {\textbf{Due date}} & {28/02/2010} & {\textbf{Submitted}} & {27/02/2010}\\ \hline
	\textcolor{white}{\textbf{Status}} & \multicolumn{2}{l|}{} & \multicolumn{2}{l|}{final}\\ \hline
	\textcolor{white}{\textbf{Dissemination Level}} & \multicolumn{4}{l|}{Public $\boxtimes$ / Consortium $\square$}\\ \hline
	\textcolor{white}{\textbf{Short description of contents}} & \multicolumn{4}{p{10.85cm}|}{\deliverableNumber{} is the second iteration of the FAST Gadget ontology. The ontology provides a formal description of the conceptual model of FAST in general, and of the screen/screenflow architecture in particular. The document comprises a discussion of the ontology development methodology used, a number of intermediate representations which document the development process, and the final product in the form of an OWL ontology.}\\ \hline
	\textcolor{white}{\textbf{Authors}} & \multicolumn{4}{p{10.85cm}|}{\authorOne, \authorTwo, \authorThree, \authorFour}\\
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\
	\hline
	\textcolor{white}{\textbf{Deliverable Owner}} & \multicolumn{2}{l|}{\authorOne} & \textbf{email} & {knud.moeller@deri.org} \\ \cline{4-5}
	\textcolor{white}{\textbf{(Partner)}} & \multicolumn{2}{l|}{} & \textbf{phone} & {+353 91 495086} \\ \hline
	\textcolor{white}{\textbf{Keywords}} & \multicolumn{4}{p{10.85cm}|}{FAST, conceptual model, ontology, OWL, RDFS}\\ \hline
\end{tabular}
\end{small}
\newpage

@@

As the corner stone of the theoretical work undertaken in the FAST project, the definition of a conceptual model and ontology for the characterisation of complex gadgets feeds directly into the project's three main development activities: the development environment for complex gadgets (GVS), the development of gadget components and the semantic catalogue which represents the backend of the architecture. For all these different aspects of FAST, the ontology developed and presented in this deliverable provides a structured definition of their common domain of discourse. The ontology formally defines the parts which a complex gadget is made of, how the parts interrelate, what kind of metadata is available for each part, how users of the system are represented, etc.

@@

This deliverable is structured as follows: as part of the introduction, we will present the methodology we chose to adopt for the development of the FAST gadget ontology, by discussing  each of its proposed stages in detail. Following, we will conclude the introduction by elaborating more on some of the general design decisions we are following. After briefly addressing the topic of related work in Sect.~\ref{sec:related_work} --- which is mainly covered in a different deliverable from this work package, \cite{urmetzer2010fast_state_of_the_art} --- the structure of the remainder of this document follows directly from our adopted methodology: we will perform a domain analysis in Sect.~\ref{sec:domain_analysis}, covering the stages of specification, conceptualisation and integration of existing vocabularies. This is then followed by the implementation step in Sect.~\ref{sec:ontology}, which covers the ontology in its form at this moment of the project. Finishing the deliverable, we present our conclusions and outlook on years two and three of the project in Sect.~\ref{sec:conclusions}.

@@

\todo{update this section, refer to change requests in last review}

@@

\begin{itemize}
	\item Sect.~\ref{sub:goal_and_scope} has been revised to better define the scope of this deliverable with respect to other deliverables.
	\item The list of classes and properties has been moved to the appendix section at the back of the document (App.~\ref{sec:ontology_terms}), because it was felt that it would otherwise hinder the flow of the text.
	\item In response to the M12 review, a UML overview of the ontology in App.~\ref{sec:ontology_overview} now complements the list of terms.
	\item The section on our adopted development methodology (previously Sect.~1.3) has been moved to the end of the document (App.~\ref{sub:methontology_a_methodology_for_ontology_development}), but otherwise did not change.
	\item An evaluation section has been added (Sect.~\ref{sec:evaluation}).
	\item The section on design principles (previously Sect.~1.4, now Sect.~\ref{sub:general_design_decisions}) has been moved to the general evaluation section and extended to reflect research done by the authors within the past year.
	\item In the integration section (Sect.~\ref{sub:integration}) of the ontology definition, the Common Tag vocabulary has been added. It is briefly introduced, and the integrated terms are specified.
	\item The namespace abbreviation for the FAST gadget ontology changed from \texttt{fast} to \texttt{fgo}, in order to distinguish from other FAST ontologies which may be added in the future.
\end{itemize}

@@

\section{Related Work} % (fold)
\label{sec:related_work}

@@

\section{Domain Analysis} % (fold)
\label{sec:domain_analysis}

@@

\subsection{Conceptualisation} % (fold)
\label{sub:conceptualisation}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[scale=.66]{images/uml_composition.pdf}
    \caption{Important composition relationships within the FAST conceptual model}
    \label{fig:uml_composition}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[scale=.66]{images/uml_screen_component.pdf}
    \caption{Aggregation of screen components}
    \label{fig:uml_screen_component}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[scale=.66]{images/uml_definition.pdf}
    \caption{Building blocks have either code or are defined declaratively}
    \label{fig:uml_definition}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[scale=.66]{images/uml_condition.pdf}
    \caption{Building blocks with pre- and post-conditions}
    \label{fig:uml_condition}
  \end{center}
\end{figure}

@@

Looking at the requirements specification document, DC can be integrated into the FAST gadget ontology in many ways. A lot of the metadata needs for screens, screenflows and their components are identical to those of the media resources originally intended by DC. For this reason, many properties such as \texttt{title}, \texttt{creator}, \texttt{rights} or various \texttt{date} properties can be used directly (more details in the integration document). 
% subsubsection dublin_core (end)
\subsubsection{FOAF} % (fold)
\label{ssub:foaf}

@@

\singlespacing
\lstset{
	caption={Common Tag example --- disambiguating the tag ``cocoa''}, 
	label=list:common_tag,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/common_tag_example.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\begin{small}
\begin{longtable}{|p{3cm}|p{1.5cm}|p{3.5cm}|p{6.1cm}|}
\caption{\label{tab:integration_document}FAST Ontology integration document}\\
\hline
\rowcolor{fast@lightgrey}\textcolor{white}{\textbf{FAST Term}} & 
\textcolor{white}{\textbf{Integrated Ontology}} & 
\textcolor{white}{\textbf{Integrated Term}} & 
\textcolor{white}{\textbf{Comment}} \\ \hline
\endfirsthead
\multicolumn{4}{|c|}{\emph{\textbf{Classes}}} \\ \hline
\texttt{User} & FOAF & \texttt{foaf:Person} & A user in FAST will be modelled using a combination of FOAF and SIOC, so that an instance of \texttt{foaf:Person} will have an account on the FAST platform, represented by an instance of \texttt{sioc:User}, which is a subclass of \texttt{foaf:OnlineAccount}. The relationship is expressed using \texttt{foaf:holdsAccount}. This way of modelling directly follows the suggestions made by~\cite{sioc_related2007}.\\ \hline
\texttt{User} & SIOC & \texttt{sioc:User} & s.a.\\ \hline
\texttt{Image} & FOAF & \texttt{foaf:Image} & --- \\ \hline
\texttt{Tag} & Common Tag & \texttt{ctag:Tag} & Complex or conceptual tags are being used e.g. to model the domain context of resources.	 \\ \hline
\multicolumn{4}{|c|}{\emph{\textbf{Properties}}} \\ \hline
\texttt{hasLabel} & DC & \texttt{dcterms:title} & ---  \\ \hline
\texttt{hasCreator} & DC & \texttt{dcterms:creator} & DC defines the range of \texttt{dcterms:creator} to be \texttt{dcterms:Agent}. By inference, this obviously also holds within FAST. However, we will explicitly only use \texttt{foaf:Person} (also see Fig.~\ref{fig:dcterms_example}). \\ \hline
\texttt{hasDescription} & DC & \texttt{dcterms:description} & --- \\ \hline
\texttt{hasCreationDate} & DC & \texttt{dcterms:created} & Dates should be formatted according to ISO 8601~\cite{w3c_iso8601_1997}. \\ \hline
\texttt{hasRights} & DC & \texttt{dcterms:rights} & DC terms specify a number of classes to model the rights of resources. \texttt{dcterms:rights} links a resource to a \texttt{dcterms:RightsStatement}. The holder of the rights is specified by pointing from the resource to a \texttt{dcterms:Agent}, using the \texttt{dcterms:rightsHolder} property. \\ \hline
% \texttt{hasDomainContext} & DC & \texttt{dcterms:subject} & DC specifies that subject shoulds be non-literals. To allow simple tagging, we should represent tags as resources (\todo{common tag}). \\ \hline
\texttt{hasImage} & FOAF & \texttt{foaf:depiction} & The inverse \texttt{foaf:depicts} could also be used. \\ \hline
\texttt{hasIcon} & FOAF & sub-property of \texttt{foaf:depiction} & Icons are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasScreenshot} & FOAF & sub-property of \texttt{foaf:depiction} & Screenshots are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasName} & FOAF & \texttt{foaf:name} & Apart from \texttt{foaf:name}, which is used for the full name, more fine-grained properties such as \texttt{foaf:firstName}, \texttt{foaf:surname} or \texttt{foaf:title} and \texttt{foaf:nick} can be used as well. \\ \hline
\texttt{hasEmail} & FOAF & \texttt{foaf:mbox} & While \texttt{foaf:mbox} will be used for the plain e-mail address, \texttt{foaf:mbox\_sha1sum} will be used for an obfuscated version of the address. \\ \hline
\texttt{hasInterests} & FOAF & \texttt{foaf:interest} & --- \\ \hline
\texttt{hasUserName} & FOAF & \texttt{foaf:accountName} & --- \\ \hline
\texttt{hasDomainContext} & Common Tag & \texttt{ctag:tagged} & The domain context of a resource in FAST can be expressed as a complex tag. Further Common Tag properties can then be applied to make assertions about the tag itself. \\ \hline
\texttt{hasVersion} & DOAP & \texttt{doap:revision} & DOAP (Description of a Project) is a widely used vocabulary for basic annotation of software and other projects. \\ \hline
\end{longtable}
\end{small}
\doublespacing

%% subsubsection integration_document (end)
%
%% subsection integration (end)
%
%% section domain_analysis (end)
%
\section{The FAST Gadget Ontology} % (fold)
\label{sec:ontology}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\linewidth]{images/classes_templates_instances.pdf}
    \caption{Different levels of representation of FAST resources}
    \label{fig:classes_templates}
  \end{center}
\end{figure}

@@

\singlespacing
\lstset{
	caption={Templating example --- different levels of representation of a FAST screen (TriG)}, 
	label=list:templating,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/templating_example.ttl}
\end{figure}
\doublespacing

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=.75\linewidth]{images/templating_example.pdf}
    \caption{Templating example --- different levels of representation of a FAST screen (diagramme)}
    \label{fig:templating_example_graph}
  \end{center}
\end{figure}

@@

\singlespacing
\lstset{
	caption={Example of basic annotation of a FAST resource}, 
	label=list:basic_annotation,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/screen_example.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Basic description of a user in FAST}, 
	label=list:user,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/user_example_01.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Modelling the creator of a resource}, 
	label=list:creator,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/user_example_02.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Simple pre-condition}, 
	label=ex:simple_precondition,
	language=sparql
}
\begin{figure}[ht]
\begin{lstlisting}
	?user a sioc:User .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Post-condition at the prototypical stage}, 
	label=ex:postcondition,
	language=sparql
}
\begin{figure}[ht]
\begin{lstlisting}
	?user a sioc:User ;
	      foaf:accountName ?account_name .
	?person a foaf:Person ;
	      foaf:holdsAccount ?user ;
	      foaf:name ?person_name .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Post-condition as instantiated during design time (on the FAST GVS canvas)}, 
	label=ex:postcondition_canvas,
	language=sparql
}
\begin{figure}[ht]
\begin{lstlisting}
	_:b0 a sioc:User ;
	      foaf:accountName _:b1 .
	_:b2 a foaf:Person ;
	      foaf:holdsAccount _:b0 ;
	      foaf:name _:b3 .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Post-condition during runtime}, 
	label=ex:postcondition_runtime,
	language=sparql
}
\begin{figure}[ht]
\begin{lstlisting}
	<http://fast.org/gvs/knud> a sioc:User ;
	      foaf:accountName "dunk" .
	<http://kantenwerk.org/metadata/foaf.rdf#me> a foaf:Person ;
	      foaf:holdsAccount <http://fast.org/gvs/knud> ;
	      foaf:name "Knud Moeller" .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Linking screens to conditions}, 
	label=list:post_condition,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/post_condition_example.ttl}
\end{figure}
\doublespacing

@@

		\begin{tabular}{lcc}
			\toprule
			& \textbf{pre-condition} & \textbf{post-condition} \\ 
			\midrule
			\emph{isPositive ``YES''} & condition must be fulfilled & condition instantiated\\ 
			\emph{isPositive ``NO''} & $\neg$condition must be fulfilled & instantiated condition removed \\ 
			\bottomrule
		\end{tabular}
	\end{center}
\end{table}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\linewidth]{images/declarativeDefinitionExample.pdf}
    \caption{Explicit piping within a screen from a post- to a pre-condition (diagramme)}
    \label{fig:piping}
  \end{center}
\end{figure}

@@

\singlespacing
\lstset{
	caption={Explicit piping within a screen from a post- to a pre-condition (code)}, 
	label=list:piping,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/piping_example.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Defining a screen's domain context using the Common Tag vocabulary}, 
	label=list:domain_context,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/domain_context_example.ttl}
\end{figure}
\doublespacing

@@

\section{Evaluation} % (fold)
\label{sec:evaluation}

@@

At the moment, the use case for the FAST gadget ontology is the tool-internal representation of gadgets and their components, which means that simplicity for users in the wider community is only of secondary importance. However, the ontology is still kept relatively basic, both in terms of the number of terms (currently 19 classes and 22 properties\todo{Recount!}), and in the complexity of the class definitions themselves, which are mostly restricted to a straightforward class subsumption hierarchy, with the exception of the sparse use of the boolean class expression \texttt{owl:unionOf}.

>>>>

\lstdefinelanguage{turtle} 
{morekeywords={@prefix, a, true, false}, 
sensitive=false, 
morecomment=[l]{\#}, 
morestring=[b]",
}

@@

This deliverable defines the abstract conceptual model and concrete ontology for the semantic characterisation of complex gadgets in the FAST project. As such, it feeds directly into the various implementation efforts within the project. 
Throughout the document, we apply the \emph{Methontology} ontology development methodology. Our design and development process is therefore defined by a number of different living documents, which function both as a development tool, as well as documentation for the ontology itself. The major part of the deliverable is made up of these documents, resulting in the final \emph{implementation document} which formally defines the ontology in OWL DL semantics. Throughout the document, we provide extensive code examples and visual representations of the different aspects of both model and ontology.

@@

\begin{tabular}
	{| >{\columncolor{fast@lightgrey}}p{3.25cm}|p{1.4cm}|p{3.28cm}|p{1.6cm}|p{3.29cm}|}
	\hline
	\textcolor{white}{\textbf{Delivery data}} & {\textbf{Due date}} & {28/02/2010} & {\textbf{Submitted}} & {27/02/2010}\\ \hline
	\textcolor{white}{\textbf{Status}} & \multicolumn{2}{l|}{} & \multicolumn{2}{l|}{final}\\ \hline
	\textcolor{white}{\textbf{Dissemination Level}} & \multicolumn{4}{l|}{Public $\boxtimes$ / Consortium $\square$}\\ \hline
	\textcolor{white}{\textbf{Short description of contents}} & \multicolumn{4}{p{10.85cm}|}{\deliverableNumber{} is the second iteration of the FAST Gadget ontology. The ontology provides a formal description of the conceptual model of FAST in general, and of the screen/screenflow architecture in particular. The document a number of successive intermediate representations which document the development process, and the final product in the form of an OWL ontology. Numerous code examples and visual representations of the ontology terms are provided.}\\ \hline
	\textcolor{white}{\textbf{Authors}} & \multicolumn{4}{p{10.85cm}|}{\authorOne, \authorTwo, \authorThree, \authorFour}\\
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\
	\hline
	\textcolor{white}{\textbf{Deliverable Owner}} & \multicolumn{2}{l|}{\authorOne} & \textbf{email} & {knud.moeller@deri.org} \\ \cline{4-5}
	\textcolor{white}{\textbf{(Partner)}} & \multicolumn{2}{l|}{} & \textbf{phone} & {+353 91 495086} \\ \hline
	\textcolor{white}{\textbf{Keywords}} & \multicolumn{4}{p{10.85cm}|}{FAST, conceptual model, ontology, OWL, RDFS}\\ \hline
\end{tabular}
\end{small}
\newpage

@@

As the corner stone of the theoretical work undertaken in the FAST project, the definition of a conceptual model and ontology for the characterisation of complex gadgets feeds directly into the project's three main development activities: the development environment for complex gadgets (GVS), the development of gadget components and the semantic catalogue which represents the backend of the architecture. For all these different aspects of FAST, the ontology developed and presented in this deliverable provides a structured definition of their common domain of discourse. The ontology formally defines the parts which a complex gadget is made of, how the parts interrelate, what kind of metadata is available for each part, how users of the system are represented, etc. It should be pointed out that the conceptual model and ontology are not meant to cover every conceivable aspect of the development of complex gadgets, not even every conceivable aspect of the FAST platform. Rather, the goal is the following: FAST is about the \emph{development of complex gadgets for end users}, and so the conceptual model and ontology provide the terminology and conceptual framework that is needed for specifically this task. In other words, while this document provides the theoretical underpinning for the implementation-oriented work in the project, the implementation-oriented work on the other hand defines the scope and requirements for the model.

@@

This deliverable is structured as follows: 
we will start off by outlining the main changes that this document has undergone since its previous version, so as to enable the reader to quickly identify new material.
After briefly addressing the topic of related work in Sect.~\ref{sec:related_work} --- which is mainly covered in a different deliverable from this work package, \cite{urmetzer2010fast_state_of_the_art} --- the structure of the following sections follows directly from our adopted ontology development methodology: we will perform a domain analysis in Sect.~\ref{sec:domain_analysis}, covering the stages of specification, conceptualisation and integration of existing vocabularies. This is then followed by the implementation step in Sect.~\ref{sec:ontology}, which covers the concrete ontology in its form at this moment of the project. 
After a section discussing an evaluation of the FAST gadget ontology, the deliverable will be concluded with a summary of our work in Sect.~\ref{sec:conclusions}.\ref{sub:methontology_a_methodology_for_ontology_development}
A number of appendices have been added to the document: App.~\ref{sec:methodology} presents the development methodology adopted by us, App.~\ref{sec:ontology_terms} provides documentation for the ontology in the form of a detailed list of terms, while App.~\ref{sec:ontology_code} contains the complete code of the ontology.

@@

@@

\begin{itemize}
	\item Sect.~\ref{sub:goal_and_scope} has been revised to better define the scope of this deliverable with respect to other deliverables.
	\item The list of classes and properties has been moved to the appendix section at the back of the document (App.~\ref{sec:ontology_terms}), because it was felt that it would otherwise hinder the flow of the text.
	\item In response to the M12 review, numerous graphical representations both of the underlying conceptual model as well as the concrete gadget ontology have been added to the document, using UML notation. As an example, Sect.~\ref{sub:ontology_overview} now provides a complete overview of the ontology with all classes and properties.
	\item Also in response to the previous review, conceptual model and ontology have been extended both in breadth and depth.
	\item As an addition to the glossary of terms in the conceptualisation phase, a section discussing and illustrating the relations between the different classes has been provided.
	\item In order to reflect the extension of the scope of FAST towards screen creation (as opposed to screen flow creation), the conceptual model and ontology have been extended accordingly.
	\item A mechanism for RDF templating has been devised and is discussed in Sect.~\ref{sub:classes_instances_and_templates}.
	\item As the result of the ongoing dialogue between ontology developers and application implementers, many classes and properties have been added, changed or removed.
	\item The section on our adopted development methodology (previously Sect.~1.3) has been moved to the end of the document (App.~\ref{sub:methontology_a_methodology_for_ontology_development}), but otherwise did not change significantly.
	\item An evaluation section has been added (Sect.~\ref{sec:evaluation}), based on research done by the authors within the past year.
	\item The section on design principles (previously Sect.~1.4, now Sect.~\ref{sub:general_design_decisions}) has been moved to the general evaluation section.
	\item In the integration section (Sect.~\ref{sub:integration}) of the ontology definition, the Common Tag vocabulary for complex tagging has been added. It is briefly introduced, and the integrated terms are specified.
	\item On a technical note, the namespace abbreviation for the FAST gadget ontology changed from \texttt{fast} to \texttt{fgo}, in order to distinguish from other FAST ontologies which may be added in the future.
\end{itemize}

@@

\clearpage
\section{Related Work} % (fold)
\label{sec:related_work}

@@

\clearpage
\section{Domain Analysis and Conceptualisation} % (fold)
\label{sec:domain_analysis}

@@

\subsection{Conceptual Model} % (fold)
\label{sub:conceptualisation}

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=.66]{images/uml_composition.pdf}
    \caption{Important composition relationships within the FAST conceptual model}
    \label{fig:uml_composition}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=.66]{images/uml_screen_component.pdf}
    \caption{Aggregation of screen components}
    \label{fig:uml_screen_component}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=.66]{images/uml_definition.pdf}
    \caption{Building blocks have either code or are defined declaratively}
    \label{fig:uml_definition}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=.66]{images/uml_condition.pdf}
    \caption{Building blocks with pre- and post-conditions}
    \label{fig:uml_condition}
  \end{center}
\end{figure}

@@

Looking at the requirements specification document, DC can be integrated into the FAST gadget ontology in many ways. A lot of the metadata needs for screens, screenflows and their components are identical to those of the media resources originally intended by DC. For this reason, many properties such as \texttt{title}, \texttt{creator}, \texttt{rights} or various \texttt{date} properties can be used directly (more details in the integration document). 
% subsubsection dublin_core (end)

\subsubsection{FOAF} % (fold)
\label{ssub:foaf}

@@

\singlespacing
\lstset{
	caption={Common Tag example --- disambiguating the tag ``cocoa''}, 
	label=list:common_tag,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/common_tag_example.ttl}
\end{figure}
\doublespacing

@@

\clearpage
\singlespacing
\begin{small}
\begin{longtable}{|p{3cm}|p{1.5cm}|p{3.5cm}|p{6.1cm}|}
\caption{\label{tab:integration_document}FAST Ontology integration document}\\
\hline
\rowcolor{fast@lightgrey}\textcolor{white}{\textbf{FAST Term}} & 
\textcolor{white}{\textbf{Integrated Ontology}} & 
\textcolor{white}{\textbf{Integrated Term}} & 
\textcolor{white}{\textbf{Comment}} \\ \hline
\endfirsthead
\multicolumn{4}{|c|}{\emph{\textbf{Classes}}} \\ \hline
\texttt{User} & FOAF & \texttt{foaf:Person} & A user in FAST will be modelled using a combination of FOAF and SIOC, so that an instance of \texttt{foaf:Person} will have an account on the FAST platform, represented by an instance of \texttt{sioc:User}, which is a subclass of \texttt{foaf:OnlineAccount}. The relationship is expressed using \texttt{foaf:holdsAccount}. This way of modelling directly follows the suggestions made by~\cite{sioc_related2007}.\\ \hline
\texttt{User} & SIOC & \texttt{sioc:User} & s.a.\\ \hline
\texttt{Image} & FOAF & \texttt{foaf:Image} & --- \\ \hline
\texttt{Tag} & Common Tag & \texttt{ctag:Tag} & Complex or conceptual tags are being used e.g. to model the domain context of resources.	 \\ \hline
\multicolumn{4}{|c|}{\emph{\textbf{Properties}}} \\ \hline
\texttt{hasLabel} & DC & \texttt{dcterms:title} & ---  \\ \hline
\texttt{hasCreator} & DC & \texttt{dcterms:creator} & DC defines the range of \texttt{dcterms:creator} to be \texttt{dcterms:Agent}. By inference, this obviously also holds within FAST. However, we will explicitly only use \texttt{foaf:Person} (also see Fig.~\ref{fig:dcterms_example}). \\ \hline
\texttt{hasDescription} & DC & \texttt{dcterms:description} & --- \\ \hline
\texttt{hasCreationDate} & DC & \texttt{dcterms:created} & Dates should be formatted according to ISO 8601~\cite{w3c_iso8601_1997}. \\ \hline
\texttt{hasRights} & DC & \texttt{dcterms:rights} & DC terms specify a number of classes to model the rights of resources. \texttt{dcterms:rights} links a resource to a \texttt{dcterms:RightsStatement}. The holder of the rights is specified by pointing from the resource to a \texttt{dcterms:Agent}, using the \texttt{dcterms:rightsHolder} property. \\ \hline
% \texttt{hasDomainContext} & DC & \texttt{dcterms:subject} & DC specifies that subject shoulds be non-literals. To allow simple tagging, we should represent tags as resources (\todo{common tag}). \\ \hline
\texttt{hasImage} & FOAF & \texttt{foaf:depiction} & The inverse \texttt{foaf:depicts} could also be used. \\ \hline
\texttt{hasIcon} & FOAF & sub-property of \texttt{foaf:depiction} & Icons are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasScreenshot} & FOAF & sub-property of \texttt{foaf:depiction} & Screenshots are a specific kind of image, so we can relate this property to the \texttt{foaf:depiciton}. \\ \hline
\texttt{hasName} & FOAF & \texttt{foaf:name} & Apart from \texttt{foaf:name}, which is used for the full name, more fine-grained properties such as \texttt{foaf:firstName}, \texttt{foaf:surname} or \texttt{foaf:title} and \texttt{foaf:nick} can be used as well. \\ \hline
\texttt{hasEmail} & FOAF & \texttt{foaf:mbox} & While \texttt{foaf:mbox} will be used for the plain e-mail address, \texttt{foaf:mbox\_sha1sum} will be used for an obfuscated version of the address. \\ \hline
\texttt{hasInterests} & FOAF & \texttt{foaf:interest} & --- \\ \hline
\texttt{hasUserName} & FOAF & \texttt{foaf:accountName} & --- \\ \hline
\texttt{hasDomainContext} & Common Tag & \texttt{ctag:tagged} & The domain context of a resource in FAST can be expressed as a complex tag. Further Common Tag properties can then be applied to make assertions about the tag itself. \\ \hline
\texttt{hasVersion} & DOAP & \texttt{doap:revision} & DOAP (Description of a Project) is a widely used vocabulary for basic annotation of software and other projects. \\ \hline
\end{longtable}
\end{small}
\doublespacing

%% subsubsection integration_document (end)
%
%% subsection integration (end)
%
%% section domain_analysis (end)
%
\clearpage
\section{The FAST Gadget Ontology} % (fold)
\label{sec:ontology}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{images/classes_templates_instances.pdf}
    \caption{Different levels of representation of FAST resources}
    \label{fig:classes_templates}
  \end{center}
\end{figure}

@@

\singlespacing
\lstset{
	caption={Templating example --- different levels of representation of a FAST screen (code)}, 
	label=list:templating,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/templating_example.ttl}
\end{figure}
\doublespacing

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=.75\linewidth]{images/templating_example.pdf}
    \caption{Templating example --- different levels of representation of a FAST screen (diagramme)}
    \label{fig:templating_example_graph}
  \end{center}
\end{figure}

@@

\singlespacing
\lstset{
	caption={Example of basic annotation of a FAST resource}, 
	label=list:basic_annotation,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/screen_example.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Basic description of a user in FAST}, 
	label=list:user,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/user_example_01.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Modelling the creator of a resource}, 
	label=list:creator,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/user_example_02.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Simple pre-condition}, 
	label=ex:simple_precondition,
	language=sparql
}
\begin{figure}
\begin{lstlisting}
	?user a sioc:User .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Post-condition at the prototypical stage}, 
	label=ex:postcondition,
	language=sparql
}
\begin{figure}
\begin{lstlisting}
	?user a sioc:User ;
	      foaf:accountName ?account_name .
	?person a foaf:Person ;
	      foaf:holdsAccount ?user ;
	      foaf:name ?person_name .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Post-condition as instantiated during design time (on the FAST GVS canvas)}, 
	label=ex:postcondition_canvas,
	language=sparql
}
\begin{figure}
\begin{lstlisting}
	_:b0 a sioc:User ;
	      foaf:accountName _:b1 .
	_:b2 a foaf:Person ;
	      foaf:holdsAccount _:b0 ;
	      foaf:name _:b3 .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Post-condition during runtime}, 
	label=ex:postcondition_runtime,
	language=sparql
}
\begin{figure}
\begin{lstlisting}
	<http://fast.org/gvs/knud> a sioc:User ;
	      foaf:accountName "dunk" .
	<http://kantenwerk.org/metadata/foaf.rdf#me> a foaf:Person ;
	      foaf:holdsAccount <http://fast.org/gvs/knud> ;
	      foaf:name "Knud Moeller" .
\end{lstlisting}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Linking screens to conditions}, 
	label=list:post_condition,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/post_condition_example.ttl}
\end{figure}
\doublespacing

@@

		\begin{tabular}{lcc}
			\toprule
			& \textbf{pre-condition} & \textbf{post-condition} \\ 
			\midrule
			\emph{isPositive: true} & condition must be fulfilled & condition instantiated\\ 
			\emph{isPositive: false} & $\neg$condition must be fulfilled & instantiated condition removed \\ 
			\bottomrule
		\end{tabular}
	\end{center}
\end{table}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{images/declarativeDefinitionExample.pdf}
    \caption{Explicit piping within a screen from a post- to a pre-condition (diagramme)}
    \label{fig:piping}
  \end{center}
\end{figure}

@@

\singlespacing
\lstset{
	caption={Explicit piping within a screen from a post- to a pre-condition (code)}, 
	label=list:piping,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/piping_example.ttl}
\end{figure}
\doublespacing

@@

\singlespacing
\lstset{
	caption={Defining a screen's domain context using the Common Tag vocabulary}, 
	label=list:domain_context,
	language=turtle
}
\begin{figure}
	\lstinputlisting{exampleCode/domain_context_example.ttl}
\end{figure}
\doublespacing

@@

\clearpage
\section{Evaluation} % (fold)
\label{sec:evaluation}

@@

At the moment, the use case for the FAST gadget ontology is the tool-internal representation of gadgets and their components, which means that simplicity for users in the wider community is only of secondary importance. However, the ontology is still kept relatively basic, both in terms of the number of terms (currently 18 classes and 22 properties), and in the complexity of the class definitions themselves, which are mostly restricted to a straightforward class subsumption hierarchy, with the exception of the sparse use of the boolean class expression \texttt{owl:unionOf}.
